%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Property Kernels}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

For a single property type (e.g. hydrophobicity) and property vector $\mathbf{p}$ corresponding to a sequence $\mathbf{s}\in\mathcal{S}$, $\mathbf{p}$ can be split into $n$-dimensional sub-vectors $\mathbf{p}_{[1,n]}$,$\mathbf{p}_{[2,n+1]}$,$\ldots$, $\mathbf{p}_{[\ell-n+1,\ell]}$, where $\mathbf{p}_{[i,i+j]}=(p_{i},p_{i+1},...,p_{i+j})$, and $n\ll\ell$ is a small integer. For example, $\mathbf{p}_{[1,n]}$ corresponds to the first $n$ elements of $\mathbf{p}$ as shown in Figure \ref{fig:VQ_schematic}(a). For a property vector (amino acid sequence) of length $\ell$, there are $\ell-n+1$ length-$n$ sub-vectors.

Given a set of $n$ length property vectors $\mathcal{P}=\{\mathbf{p}_{1},\mathbf{p}_{2},\mathbf{p}_{3},...\}$ derived from a universe of amino acid sequences $\mathcal{S}=\{\mathbf{s}_{1},\mathbf{s}_{2},\mathbf{s}_{3},...\}$, a partition of $\mathbb{R}^{n}$ into $m$ regions $\mathcal{R}=\left\{ R_{1},R_{2},\ldots,R_{m}\right\} $ is generated. Regions are represented by a set of $n$-dimensional vectors, or centroids, $\mathcal{C}=\left\{ \mathbf{c}_{1},\mathbf{c}_{2},\ldots,\mathbf{c}_{m}\right\} $. Each region $R_{i}$ represents a Voronoi region such that 

\[
R_{i}=\left\{ \mathbf{x}:d(\mathbf{x},\mathbf{c}_{i})\leq d(\mathbf{x},\mathbf{c}_{j}),j\in\left\{ 1,2,\ldots,m\right\} \right\} ,
\]
where $d(\mathbf{x},\mathbf{c})$ is the Euclidean distance between vector $\mathbf{x}$ and centroid $\mathbf{c}$. Set $\mathcal{C}$ is determined using the k-means clustering algorithm, where the initial set of clusters is generated by the splitting method \citep{gersho1992vector}.


%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Count-Based Property Kernels}
\label{sec:propertykernels}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

Variable length property vectors can be transformed into fixed $m$-length vectors using the partition of $\mathbb{R}^{n}$ defined by $\mathcal{R}$. Specifically, a property vector $\mathbf{p}$ is mapped into a vector of length $m$ as 

\[
\mathbf{x}=\left(\varphi_{1}(\mathbf{p}),\varphi_{2}(\mathbf{p}),\ldots,\varphi_{m}(\mathbf{p})\right),
\]
where $\varphi_{i}(\mathbf{p})$ is the number of $n$-dimensional vectors $\mathbf{p}_{[.]}$ in $\mathbf{p}$ that belong to partition $R_{i}$. Given two property vectors $\mathbf{p}$ and $\mathbf{q}$ and their respective count vectors $\mathbf{x}$ and $\mathbf{y}$, a \textit{count-based property kernel} is defined as

\[
k(\mathbf{p},\mathbf{q})=\mathbf{x}^{T}\mathbf{y},
\]
where $T$ is the transpose operator. Note that in this notation each property vector is assumed to be a column vector, i.e. $\left(a,b,c\right)=\left[a\; b\; c\right]^{T}$, as in \citet{strang2003introduction}. Since the function $k(\mathbf{p},\mathbf{q})$ is defined as an inner product between two count vectors, it is a kernel function \citep{Haussler1999}.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Similarity augmented count based property kernel}
\label{sec:distancekernel}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

The count-based property kernels can be generalized by utilizing similarities between centroids to augment partial counts to each of the regions in $\mathcal{R}$. First, a
similarity function between any two centroids is defined as 

\[
s(\mathbf{c}_{i},\mathbf{c}_{j})=e^{-\alpha\cdot\frac{d(\mathbf{c}_{i},\mathbf{c}_{j})}{d_{max}}},
\]
where $d(\mathbf{c}_{i},\mathbf{c}_{j})$ is the Euclidean distance between $\mathbf{c}_{i}$ and $\mathbf{c}_{j}$ and $d_{max}$ is the maximum distance between any two centroids in $\mathcal{C}$. The parameter $\alpha$ is a positive constant that defines an extent to which large distances are penalized by the non-linear mapping above, and is optimized in the model evaluation step. A generalized property kernel can now be defined as

\[
k_{g}(\mathbf{p},\mathbf{q})=(\mathbf{Sx})^{T}(\mathbf{Sy})=\mathbf{x}^{T}\mathbf{S}^{T}\mathbf{S}\mathbf{y}=\mathbf{x}^{T}\mathbf{Q}\mathbf{y},
\]
where elements of the similarity matrix $\mathbf{S}$ are defined as $s_{ij}=s(\mathbf{c}_{i},\mathbf{c}_{j})$. Clearly, for $\mathbf{S}=\mathbf{I}$, i.e. $\alpha\rightarrow\infty$, the generalized kernel is equivalent to the basic count-based property kernel. If \textbf{$\mathbf{S}$} is a non-singular matrix, it is easy to show that $\mathbf{Q}$ is a positive definite matrix \citep{strang2003introduction}. Therefore, $k_{g}(\mathbf{p},\mathbf{q})$ is a kernel function. Note that the generalized count-based kernel does not lead to a sparse representation, even for large $m$. 


%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Property Kernels}
\label{sec:additivekernels}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

Let us again consider vector $\mathbf{p}$ and partition $\mathcal{R}$. We map $\mathbf{p}$ into an $m$-dimensional as 

\[
\mathbf{x}=\left(\psi_{1}(\mathbf{p}),\psi_{2}(\mathbf{p}),\ldots,\psi_{m}(\mathbf{p})\right),
\]
where 

\[
\psi_{i}(\mathbf{p})=\sum_{j=1}^{\ell-n+1}s\left(\mathbf{p}_{\left[j,j+n-1\right]},\mathbf{c}_{i}\right),
\]
is the cumulative similarity between vectors $\mathbf{p}_{[.]}$ in $\mathbf{p}$ and centroid $\mathbf{c}_{i}$, and $s\left(\cdot\right)$ is the vector similarity function defined above. Given two property vectors $\mathbf{p}$ and $\mathbf{q}$ and their respective count vectors $\mathbf{x}$ and $\mathbf{y}$, we now define an \textit{additive property kernel} as $k(\mathbf{p},\mathbf{q})=\mathbf{x}^{T}\mathbf{y}$. 


%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Combined Kernel Function}
\label{sec:combinedkernels}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

Let $k_{hy}(\mathbf{x},\mathbf{y})$, $k_{fl}(\mathbf{x},\mathbf{y})$, $k_{he}(\mathbf{x},\mathbf{y})$, $k_{sh}(\mathbf{x},\mathbf{y})$, $k_{id}(\mathbf{x},\mathbf{y})$, be the property kernels representing hydrophobicity, flexibility, helix propensity, sheet propensity, and intrinsic disorder, respectively. The composite property kernel is calculated as a linear combination

\[
k(\mathbf{x},\mathbf{y})=k_{hy}(\mathbf{x},\mathbf{y})+k_{fl}(\mathbf{x},\mathbf{y})+k_{he}(\mathbf{x},\mathbf{y})+k_{sh}(\mathbf{x},\mathbf{y})+k_{id}(\mathbf{x},\mathbf{y}),
\]
where prior to combining, each kernel is normalized using

\[
k(\mathbf{x},\mathbf{y})\leftarrow\frac{k(\mathbf{x},\mathbf{y})}{\sqrt{k(\mathbf{x},\mathbf{x})k(\mathbf{y},\mathbf{y})}}.
\]



%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computational Complexity}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

The computation of the count vector can be accomplished in $O(\ell mn)$ time if each $n$-dimensional vector from $\mathbf{p}$ is compared with all centroids in $\mathcal{C}$. Approximation algorithms are available through a decision tree like organization of the centroids. In such a case, only $\log m$ operations are needed resulting in $O(\ell n\log m)$ time \citep{gersho1992vector}; however, there is no guarantee that the closest centroid will be found. Using a full-search scheme, the basic and generalized property kernels can be computed in $O(\ell nm)$ and $O(m^{2}+\ell nm)$ times, respectively. The storage requirements include $O(mn)$ for storing $\mathcal{C}$ and $O(m^{2})$ for storing \textbf{$\mathbf{Q}$}. The additive kernel has complexity $O(\ell nm)$.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{String kernel}
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

As a baseline method with which we could compare results from our VQ model we utilized a string kernel as described by \cite{Leslie2002} for a wide range of word sizes ($1\dots 10$).  For a given word size $l$, a $20^{l}$-length vector was created for a protein, where each value represented the number of times a potential $l$ length substring generated using the 20 amino acid alphabet occurred.  An $\ell$-length sequence, $\mathbf{s}$, would contain $\ell-l+1$ such overlapping strings.



