\chapter{Information-theoretic evaluation of predicted ontological annotations}
\label{ch:rumi}

%\begin{center}
%\[ pub placeholder \]
%\[ pub placeholder \]
%
%\end{center}


%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%


Ontological representations have been widely used in biomedical sciences to standardize knowledge representation and exchange \citep{Robinson2011}. Modern ontologies are typically viewed as graphs in which vertices represent terms or concepts in the domain of interest and edges represent relational ties between terms (e.g. is-a, part-of). While, in theory, there are no restrictions on the types of graphs used to implement ontologies, hierarchical organizations, such as trees or directed acyclic graphs, have been frequently used in the systematization of biological experiments, organismal phenotypes, or structural and functional descriptions of biological macromolecules.

In molecular biology, one of the most frequently used ontologies is the Gene Ontology \citep{Ashburner2000}, which standardizes the functional annotation of genes and gene products. The development of the Gene Ontology (GO) was based on the premise that the genomes of all living organisms are composed of genes whose products perform functions derived from a finite molecular repertoire. In addition to knowledge representation, GO has also facilitated large-scale analyses and automated annotation of gene product function \citep{Radivojac2013}. As the rate of accumulation of uncharacterized sequences far outpaces the rate at which biological experiments can be carried out to characterize those sequences, computational function prediction has become increasingly useful for the global characterization of genomes and proteomes as well as for guiding biological experiments via prioritization  \citep{Sharan2007,Rentzsch2009}.

The growing importance of tools for the prediction of GO annotations, especially for proteins, presents the problem of how to accurately evaluate such tools. Because terms can automatically be associated with their ancestors in the GO graph, the task of an evaluation procedure can be framed as comparing the predicted graph with the graph represented by the true, experimentally verified annotations. It should be explicitly noted that the structure of the ontology introduces dependence between terms, dependencies that must be appropriately considered when comparing two graphs. This is obviously true for terms that share an ancestor/descendant relationship, but also for terms that are not related through direct ancestry.  For example, although the terms ``H3 histone acetyltransferase activity" and ''H4 histone acetyltransferase activity" do not share a ancestor/descendant relationship and describe different behaviors of histone acetyltransferases, they nonetheless are not independent. 

Protein function is also complex and context dependent. A single biological experiment rarely results in complete characterization of a protein's function. Furthermore, annotations might not always be meaningful; a point that is particularly evident in cases where only high-throughput experiments are used for functional characterization, leading to shallow annotation graphs. These types of annotations pose a problem in evaluation as the ground truth is incomplete and noisy. 

Furthermore, different computational models produce different outputs that must be accounted for. Some models simply assign binary yes-or-no values to terms; while others might assign a different score to potentially each node in the ontology, with an expectation that a good decision threshold would be applied to provide useful annotations.
Finally, a complicating factor is posed by the fact that GO, as most current ontologies, is generally unfinished and contains a range of specificities of functional descriptions at the same depth of the ontology \citep{alterovitz2010}.

There are two important factors related to the development of evaluation metrics. First, because both the experimental and predicted annotation of genes can be represented as subgraphs of the generally much larger GO graph, it is unlikely that a given computational method will provide an exact prediction of the experimental annotation. It is therefore necessary to develop metrics that facilitate calculating degrees of similarity between pairs of graphs and appropriately address dependency between nodes. Ideally, such a measure of similarity would be able to characterize both the level of correct prediction of the true (albeit incomplete) annotation but also the level of misannotation. The second important factor related to the evaluation metric is its interpretability. Characterizing the predictor's performance should be meaningful to a downstream user. Ideally, an evaluation metric would have a simple probabilistic interpretation. 

In this chapter, we develop an information-theoretic framework for evaluating the prediction accuracy of computer-generated ontological annotations. We first use the structure of the ontology to probabilistically model, via a Bayesian network, the prior distribution of protein experimental annotation. We then apply our metric to three protein function prediction algorithms selected to highlight the limitations of typically-considered evaluation metrics. We show that our metrics provide added value to the current analyses of the strengths and weaknesses of computational tools. Finally, we argue that our framework is probabilistically well founded and show that it can also be used to augment already existing evaluation metrics.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
 

\section{Background}
The issue of performance evaluation is closely related to the problems of measuring similarity between pairs of graphs or sets. First, we note that a protein's annotation (experimental or predicted) is a graph containing a subset of nodes in the ontology together with edges connecting them. We use the term \emph{leaf node} to describe a node that has no descendants in the annotation graph, although it is allowed to have descendants in the ontology. A set of leaf terms completely describes the annotation graph. %In the literature, such nodes are often referred to simply as terms with an assumption that all of their ancestors are automatically included in the annotation of the protein.


We roughly group both graph similarity and performance evaluation metrics into topological and probabilistic categories, and note that a particular metric may combine aspects from both. More elaborate distinctions are provided by \cite{Pesquita2009} and \cite{Guzzi2012}. Topological metrics rely on the structure of the ontology to perform evaluation and typically employ metrics that operate on sets of nodes and/or edges. A number of topological measures have been utilized, including the Jaccard and cosine similarity coefficients (the cosine function maps the binary term designations into a vector space), the shortest path-based distances \citep{Rada1989}, etc. In the context of classifier performance analysis, two common 2D metrics are the precision/recall curve and the Receiver Operating Characteristic (ROC) curve. Both curves are constructed based on the overlap in either edges or nodes between true and predicted terms and have been widely used to evaluate the performance of tools for the inference of GO annotations. They can also be used to provide a single statistic to rank classifiers through the maximum F-measure in the case of precision/recall curve or the area under the ROC curve. The area under the ROC curve has a limitation arising from the fact that the ontology is relatively large, but that the number of terms associated with a typical protein is relatively small. In practice, this results in specificities close to one regardless of the prediction, as long as the number of predicted terms is relatively small.

Although these statistics provide good feedback regarding multiple aspects of a predictor's performance, they do not always address node dependency or the problem of unequal specificity of functional annotations found at the same depth of the graph. Coupled with a large bias in the distribution of terms among proteins, prediction methods that simply learn the prior distribution of terms in the ontology could appear to have better performance than they actually do.

The second class of similarity/performance measures are probabilistic or information-theoretic metrics. Such measures assume an underlying probabilistic model over the ontology and use a database of proteins to learn the model. Similarity is then assessed by measuring the information content of the shared terms in the ontology but can also take into account the information content of the individual annotations. Unlike with topological measures where updates to the ontology  affect similarity between objects, information theoretic measures are also affected by changes in the underlying probabilistic model even if the structure of the ontology remains the same.

Probabilistic metrics closely follow and extend the methodology laid out by \cite{Resnik1995}, which is based on the notion of information content between a pair of individual terms. These measures overcome biases related to the structure of the ontology; however, they have several drawbacks of their own. One that is especially important in the context of analyzing the performance of a predictor is that they only report a single statistic, namely the similarity or distance between two terms or sets of terms. This ignores the tradeoff between precision and recall that any predictor has to make. In the case of Resnik's metric, a prediction by any descendant of the true term will be scored as if it is an exact prediction. Similarly, a shallow prediction will be scored the same as a prediction that deviates from the true path at the same point, regardless of how deep the erroneous prediction might be. Although some of these weaknesses have been corrected in subsequent work \citep{Jiang1997,Lin1998,Schlicker2006}, there remains the issue that the available probabilistic measures of semantic similarity resort to ad hoc solutions to address the common situation where proteins are annotated by graphs that contain multiple leaf terms \citep{Clark2011}. Various approaches have been taken, including averaging between all pairs of leaf terms \citep{Lord2003}, finding the maximum among all pairs \citep{Resnik1999}, or finding the best-match average, but each such solution lacks strong justification in general. For example, all-pair averaging leads to anomalies where the exact prediction of an annotation containing a single leaf term $u$ would be scored higher than the exact prediction of an annotation containing two distinct leaf terms $u$ and $v$ of equal information content, when it is more natural to think that the latter prediction should be scored higher. Finally, all semantic similarity metrics that incorporate some form of pairwise matching between leaf terms tacitly assume that the objects to be compared are annotated by similar numbers of leaf terms. As such, they could produce undesirable solutions when applied to a wide range of prediction algorithms such as those outputting a very large number of predicted terms.


%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%


\section{Protein function prediction scenarios}

In protein function prediction, there are two scenarios in which a computational model can be constructed \citep{Radivojac2013}: (i) given a new protein, the task of a classifier is to find all functional terms that the protein is associated with (``what is the function of this protein?"); and (ii) given a functional term, the task of a classifier is to find all proteins associated with this term (``what are the proteins associated with this function?"). Undoubtably, the two prediction scenarios are related because a perfect predictor of protein function would solve both questions at the same time. However, imperfect predictors may address one question better than the other. For example, a classifier built to address the first question is expected to be accurate in predicting all (or many) functional terms and that prediction scores over all functional terms are comparable. On the other hand, a predictor developed to consider only one functional term at a time need not consider any other term. Such models can perform well even if they just rank all test proteins. In addition to these differences, the two types of models are evaluated in different ways (yet they use the same terminology).

Models that are concerned with predicting function on a previously unseen protein (scenario 1, above) need to devise evaluation metrics to estimate the expected accuracy of a predicted consistent graph $P$ when the experimental (true) annotation of the protein is graph $T$. Alternatively, the models that are concerned with ranking the proteins according to their likelihood to be associated with a particular functional term $v$ (scenario 2, above) need to be evaluated based on the expectation that a particular protein is associated with a functional term $v$. Here, the models are usually evaluated for each functional term $v$, one at a time. Evaluation metrics corresponding to the former problem are significantly more challenging than the metrics corresponding to the latter problem. In the latter case, one can simply consider a particular decision threshold for predicting whether a protein is associated with function $v$ and then calculate the fraction of positive predictions that are correct (precision) as well as the fraction of proteins known to be associated with functional term $v$ that have been retrieved (recall). Such evaluation has been discussed by \citet{Sharan2007}, among others. Calculating precision and recall for the former scenario is the topic of our study.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\input{rumi_methods}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

\input{rumi_confusion}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%


\input{sdproof}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%


\input{rumi_annotation_models}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

\input{rumi_results}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%



\section{Discussion}

Here we introduce an information-theoretic framework for evaluating the performance of computational protein function prediction. We frame protein function prediction as a structured-output learning problem in which the output space is represented by consistent subgraphs of the GO graph. We argue that our approach directly addresses evaluation in cases where there are multiple true and predicted (leaf) terms associated with a protein by taking the structure of the ontology and the dependencies between terms induced by a hierarchical ontology into account. Our method also facilitates accounting for the high level of biased and incomplete experimental annotations of proteins by allowing for the weighting of proteins based on the information content of their annotations. Because we maintain an information-theoretic foundation, our approach is relatively immune to the potential dissociation between the depth of a term and its information content; a weakness of often-used topological metrics in this domain such as precision/recall or ROC based evaluation. At the same time, because we take a holistic approach to considering a protein's potentially large set of true or predicted functional associations we resolve many of the problems introduced by the practice of aggregating multiple pairwise similarity comparisons common to existing semantic similarity measures.

Although there is a long history \citep{Resnik1999} and a significant body of work in the literature regarding the use of semantic similarity measures \citep{Pesquita2009,Guzzi2012}, to the best of our knowledge all such metrics are based on single statistics and are unable to provide insight into the levels of remaining uncertainty and misinformation that every predictor is expected to balance. Therefore, the methods proposed in this work extend, modify, and formalize several useful information-theoretic metrics introduced over the past decades. In addition, both remaining uncertainty and misinformation have natural information-theoretic interpretations and can provide meaningful information to the users of computational tools. At the same time, the semantic distance based on these concepts facilitates not only the use of a single performance measure to evaluate and rank predictors, but can also be exploited as a loss function during training.

One limitation of the proposed approach is grounded in the assumption that a Bayesian network, structured according to the underlying ontology, will perfectly model the prior probability distribution of a target variable. An interesting anomaly with this approach is that the marginal probability, and subsequently the information content, of a single term (i.e. consistent graph with a single leaf term) calculated by considering the conditional probabilities between child and parent terms in the graph does not necessarily match the relative term frequency in the database. \emph{Ad hoc} solutions that maintain the term information content are possible but would result in sacrificed interpretability of the metric itself. One such solution can be obtained via a recursive definition 
\[
ia(v)=i(v)-\sum_{u\in\mathcal{P}(v)}ia(u) 
\]
where $i(v)$ is estimated directly from the database and $ia(root)=0$.

Finally, rationalizing between evaluation metrics is a difficult task. The literature presents several strategies where protein sequence similarity, protein-protein interactions or other data are used to assess whether a performance metric behaves according to expectations \citep{Guzzi2012}. In this work, we took a somewhat different approach and showed that the demonstrably biased protein function data can be shown to provide surprising results with well-understood prediction algorithms and conventional evaluation metrics. Thus, we believe that our experiments provide evidence of the usefulness of the new evaluation metric.


